---
title: "Missing data and imputation"
subtitle: "This is a Report Template Quarto"
author: "Tharun Teja Gandham (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

Missing data is a common concern, especially in the health care sector where datasets like Electronic Health Records (EHR) are sometimes incomplete for a variety of reasons, includes technological issues, in applicability, or information loss. If missing data is not appropriately considered, it might result in biased results, reduced reliability, and compromising conclusions in clinical epidemiology and other healthcare research [@salgado2016missing]. Results can be affected by missing values, which is the most common problem for statistical models that depend on complete datasets. Either imputation or elimination can be used to address missing data in order to eleminate these problems. Imputation is the recommended method since it lowers the possibility of loss of data and protects the integrity of the dataset [@pedersen2017missing].

Missing completely at random (MCAR) and missing not at random (MNAR) are two examples of the underlying mechanisms of missingness that heavily influence the choice of imputation technique. Imputation techniques range from simple techniques like mean substitution to more complex strategies like multiple imputation, which guesses missing values by generating multiple data sets and merging the resulting results to take uncertainty into account [@pedersen2017missing]. The ability of these highly sophisticated methods, primarily multiple imputation, to minimize bias and maintain the accuracy of investigations has made them popular in clinical research [@salgado2016missing]. The purpose of this project is to investigate and contrast different imputation methods, examining how well they handle diverse situations involving missing data and maintaining the validity of datasets in clinical epidemiology and healthcare research.


*This is my work and I want to add more work...*

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
